---
title: "Introduction to moby"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to moby}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(moby)
```

## Overview

`moby` is an R package for the analysis and visualization of passive acoustic
telemetry data, with a focus on marine environments. It provides a unified set of
tools spanning data cleaning, temporal and spatial classification, movement and
home-range analysis, residency metrics, and spatiotemporal overlap / social-network
analysis.

Most functions accept user-defined column names (e.g. `id.col`, `datetime.col`,
`station.col`). When several functions are applied to the same dataset, you can wrap the
data once with `as_moby()`: the resulting `mobyData` object carries the column mapping
(and, optionally, the coordinate reference system, tagging dates and ID groups) as
metadata, so these do not need to be repeated on every call. A `mobyData` object is still
an ordinary `data.frame`, and plain data frames remain fully supported.

```{r moby-object}
df <- data.frame(
  ID = c("A", "A", "B"),
  datetime = as.POSIXct(c("2023-06-01 00:00", "2023-06-01 01:00", "2023-06-01 00:00"),
                        tz = "UTC"),
  station = c("R1", "R2", "R1")
)
md <- as_moby(df)
md
```

## A small example dataset

The functions below operate on ordinary data frames, making it easy to integrate
`moby` into existing workflows. Here we build a tiny synthetic detection dataset.

```{r data}
set.seed(1)
detections <- data.frame(
  ID = factor(rep(c("A", "B"), each = 24)),
  datetime = rep(seq(as.POSIXct("2023-06-01 00:00:00", tz = "UTC"),
                     by = "1 hour", length.out = 24), 2),
  station = sample(c("R1", "R2", "R3"), 48, replace = TRUE)
)
head(detections)
```

## Temporal classification

`getTimeBins()` rounds timestamps to a regular interval, while `getSeason()` and
`getDielPhase()` assign seasonal and diel classes.

```{r temporal}
# assign 2-hour time bins
detections$timebin <- getTimeBins(detections$datetime, interval = "2 hours")

# meteorological season (northern hemisphere)
detections$season <- getSeason(detections$datetime, hemisphere = "Northern")

table(detections$season)
```

## Reshaping to wide format

`createWideTable()` converts the long-format detections into a time-bin x individual
matrix, the structure used by the overlap and visualization functions.

```{r wide}
wide <- createWideTable(detections, value.col = "detections", verbose = FALSE)
dim(wide)
```

## Where to go next

See `help(package = "moby")` for the complete list of functions, including
`filterDetections()`, `calculateStepDistances()`, `calculateUDs()`, `calculateAssociations()`,
and the family of `plot*()` visualization tools.
