---
title: "00 - Getting real data from ETN (appendix)"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

::: {.callout-important}
## Optional appendix
The core tutorials all run on the bundled `rays` dataset, so you **do not need** this module to
follow the course. It is here for when you want to pull *your own* detections from a live database —
the [European Tracking Network (ETN)](https://www.lifewatch.be/etn/) in this example. Every code
chunk here requires database credentials, so none of them run automatically.
:::

## Learning objectives

1. Install and authenticate the `etn` package.
2. Download animal metadata, deployments and detections for a project.
3. Save a detections file ready for [module 01](01-import.html).

## 1. Install required packages

```{r}
# run once
pak::pak("inbo/etn")
install.packages("usethis")
```

## 2. Store ETN credentials locally

Storing credentials in your `.Renviron` avoids typing them into every call (and into scripts you
might share).

```{r}
usethis::edit_r_environ()
```

Add these lines, save, then restart R (*Session → Restart R*):

```
ETN_USER="your_email@example.com"
ETN_PWD="your_password"
```

## 3. Explore available projects

```{r}
library(etn)
etn_projects <- etn::get_animal_projects()
View(etn_projects)
```

## 4. Download metadata, deployments and detections

```{r}
id_metadata <- etn::get_animals(animal_project_code = "Inforbiomares")

id_deployments <- etn::get_acoustic_deployments(acoustic_project_code = "Inforbiomares")

all_detections <- etn::get_acoustic_detections(
  animal_project_code = "Inforbiomares",
  scientific_name     = c("Dasyatis pastinaca", "Raja clavata")
)
all_detections <- as.data.frame(all_detections)

# record the download time for reproducibility
attr(all_detections, "download.date") <- Sys.time()
```

## 5. Save for the import tutorial

```{r}
write.csv2(all_detections, "INFORBIOMARES_detections.csv", row.names = FALSE)
```

➡️ Continue with [01 — From raw detections to a mobyData object](01-import.html), pointing it at the
file you just saved.
