---
title: "01 - From raw detections to a mobyData object"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

::: {.callout-note}
## In this module
Read raw detections from a file, harmonise them into moby's expected layout, attach tag metadata,
and wrap everything into a **`mobyData`** object that downstream functions understand. You will also
define the **`id.groups`** object that drives per-species analyses throughout the tutorials.

**Prerequisites:** none. Uses the bundled CSV at
`system.file("extdata", "rays_detections.csv", package = "moby")`. (For pulling data from a live
database, see the optional [ETN appendix](00-etn-import.html).)
:::

## Learning objectives

1. Read a generic detection export and map its columns with `importDetections()`.
2. Attach tag metadata with `importTags()` / `assignAnimalIDs()`.
3. Understand the `mobyData` data model and create one with `as_moby()`.
4. Define `id.groups` (two species) once, for reuse everywhere.

## Setup

```{r}
#| message: false
library(moby)
detections_csv <- system.file("extdata", "rays_detections.csv", package = "moby")
```

## 1. Read and harmonise the detections

The raw file uses source-specific column names (`animal_id`, `timestamp`, `station_name`, …). The
generic importer maps them onto moby's canonical names.

```{r}
#| eval: false
raw <- read.csv(detections_csv, stringsAsFactors = FALSE)
raw$timestamp <- as.POSIXct(raw$timestamp, tz = "UTC")

detections <- importDetections(
  raw,
  source = "generic",
  col.map = list(
    ID          = "animal_id",
    datetime    = "timestamp",
    station     = "station_name",
    lon         = "deploy_longitude",
    lat         = "deploy_latitude",
    transmitter = "transmitter"      # kept so assignAnimalIDs() can join tag metadata
  )
)
```

::: {.callout-tip}
moby also ships dedicated importers for common sources — `importDetections(source = "vue")`,
`"vdat"`, `"glatos"`, `"otn"`, `"etn"` — so you usually do **not** need to hand-map columns.
:::

## 2. Attach tag / animal metadata

```{r}
#| eval: false
tags <- importTags(rays_tags, source = "generic",
                   col.map = list(ID = "ID", tagging_date = "tagging_date"))

detections <- assignAnimalIDs(detections, tags)
```

## 3. Build the `mobyData` object

A `mobyData` *is* a data frame, but it carries metadata (column mapping, tagging dates, `id.groups`)
as an attribute, so you don't repeat those arguments on every call.

```{r}
#| eval: false
# one named list element per species -> drives per-species outputs everywhere
id_groups <- split(rays_tags$ID, rays_tags$species)

dataset <- as_moby(
  detections,
  id.col        = "ID",
  datetime.col  = "datetime",
  station.col   = "station",
  lon.col       = "lon",
  lat.col       = "lat",
  tagging.dates = setNames(rays_tags$tagging_date, rays_tags$ID),
  id.groups     = id_groups
)

mobyMeta(dataset)   # inspect the attached metadata
```

## Recap & what's next

You produced a `mobyData` object and defined `id.groups`. The bundled `rays` object is exactly this,
already cleaned — later modules load it directly.

➡️ **Next:** [02 — Quality control, deployment matching & filtering](02-quality-control.html).
