---
title: "01 - From raw detections to a mobyData object"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

::: {.callout-note}
## In this module
Read raw detections from a file, harmonise them into moby's expected layout, attach tag metadata,
and wrap everything into a **`mobyData`** object that downstream functions understand. You will also
define the **`id.groups`** object that drives per-species analyses throughout the tutorials.

**Prerequisites:** none. Uses the bundled CSV at
`system.file("extdata", "rays_detections.csv", package = "moby")`. (For pulling data from a live
database, see the optional [ETN appendix](00-etn-import.html).)
:::

## Learning objectives

1. Read a generic detection export and map its columns with `importDetections()`.
2. Attach tag metadata with `importTags()` / `matchTags()`.
3. Understand the `mobyData` data model and create one with `as_moby()`.
4. Define `id.groups` (two species) once, for reuse everywhere.

## Setup

```{r}
#| message: false
library(moby)
detections_csv <- system.file("extdata", "rays_detections.csv", package = "moby")
```

## 1. Read and harmonise the detections

The raw file uses source-specific column names (`animal_id`, `timestamp`, `station_name`, …). The
generic importer maps them onto moby's canonical names — and parses the date-times for you, so there
is no need to convert them beforehand.

```{r}
#| eval: false
detections <- importDetections(
  detections_csv,
  source = "generic",
  col.map = list(
    ID          = "animal_id",
    datetime    = "timestamp",
    station     = "station_name",
    lon         = "deploy_longitude",
    lat         = "deploy_latitude",
    transmitter = "transmitter"      # kept so matchTags() can join tag metadata
  )
)
```

::: {.callout-tip}
moby also ships dedicated importers for common sources — `importDetections(source = "vue")`,
`"vdat"`, `"glatos"`, `"otn"`, `"etn"` — so you usually do **not** need to hand-map columns.
See `?moby_import_schema` for every canonical field and which ones are required.
:::

::: {.callout-important}
## Importing reshapes; `as_moby()` declares
`importDetections()` returns a **plain data frame** in moby's canonical layout — it renames columns,
parses dates and derives `ID`, but attaches no study metadata. Steps 2 and 3 below turn it into a
`mobyData`. If your table is *already* tidy (your own column names, dates already parsed), skip the
importer and go straight to `as_moby()`.
:::

## 2. Attach tag / animal metadata

```{r}
#| eval: false
tags <- importTags(rays_tags, source = "generic",
                   col.map = list(ID = "ID", tagging_date = "tagging_date"))

detections <- matchTags(detections, tags)
```

`matchTags()` resolves transmitter codes to animals — several codes can belong to one animal,
typically because a sensor tag transmits on more than one code space. It changes rows and columns
only: it returns the same kind of object you gave it and attaches nothing behind your back. Its
console output names the per-animal metadata the tag table can supply — tagging dates, and
transmitter nominal delays where present — which you attach in the next step.

::: {.callout-tip}
This step is independent of `matchDeployments()` (module 02): one resolves transmitters to animals,
the other resolves receivers to deployment windows. Either order gives the same result.
:::

## 3. Build the `mobyData` object

A `mobyData` *is* a data frame, but it carries metadata (column mapping, tagging dates, `id.groups`)
as an attribute, so you don't repeat those arguments on every call.

`as_moby()` is the only function that creates one, and the single place metadata enters your data:
hand it the tag table and it derives each animal's tagging date and nominal delay, alongside what
only you can supply — here the species grouping (and, when you have them, a projected CRS and a
coastline). Because the columns already use moby's canonical names, no column arguments are needed.

```{r}
#| eval: false
# one named list element per species -> drives per-species outputs everywhere
id_groups <- split(rays_tags$ID, rays_tags$species)

dataset <- as_moby(detections, tags = tags, id.groups = id_groups)

mobyMeta(dataset)   # tagging dates, derived from 'tags' at this call
dataset             # the print method summarises what is attached
```

## Recap & what's next

You produced a `mobyData` object and defined `id.groups`. The bundled `rays` object is exactly this,
already cleaned — later modules load it directly.

➡️ **Next:** [02 — Quality control, deployment matching & filtering](02-quality-control.html).
