---
title: "02 - Quality control, deployment matching & filtering"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

::: {.callout-note}
## In this module
Make the data *trustworthy* before analysing it: audit the receiver-deployment log, match detections
to deployments, remove spurious detections, and bin detections into regular time intervals.
"Garbage in, garbage out" — this is where the science is won or lost.

**Prerequisites:** none. Uses the bundled `rays_detections` and the raw ETN deployment log shipped
at `system.file("extdata", "rays_deployments.csv", package = "moby")`.
:::

## Learning objectives

1. Import a raw receiver-deployment log with `importDeployments()`.
2. Audit deployment metadata with `checkDeployments()` and picture it with `plotDeployments()`.
3. Match detections to deployment windows with `matchDeployments()`.
4. Remove spurious detections with `filterDetections()`.
5. Bin detections into regular intervals with `getTimeBins()`.

## Setup

```{r}
#| message: false
library(moby)
data(rays_detections)
data(rays_tags)
```

## 1. Import the receiver-deployment log

Deployment logs arrive in vendor/network-specific layouts. `importDeployments()` harmonises them onto
moby's canonical schema (`receiver` / `station` / `lon` / `lat` / `deploy` / `recover`) — pick the
matching `source` preset (here the European Tracking Network export bundled with the package), or
supply a `col.map` for a format it doesn't recognise.

```{r}
#| eval: false
dep_file <- system.file("extdata", "rays_deployments.csv", package = "moby")   # raw ETN export
deployments <- importDeployments(dep_file, source = "etn")
head(deployments)
```

## 2. Audit the deployment log

`checkDeployments()` audits the receiver log and only *reports* problems — it never edits the data
or aborts — returning a structured report you can inspect or act on.

```{r}
#| eval: false
qc <- checkDeployments(deployments)
qc   # flags overlaps, impossible dates/coords, duplicates, coverage gaps
```

Supply the detections too, and it also **cross-checks them against the deployment windows** — flagging
detections from unknown receivers, detections outside any window, and station-name mismatches:

```{r}
#| eval: false
checkDeployments(deployments, detections = rays_detections, datetime.col = "datetime")
```

Every check runs by default. Use `checks =` to run only the groups you want (the coverage-gap check
is the noisiest, since routine servicing gaps are legitimate):

```{r}
#| eval: false
checkDeployments(deployments, checks = c("dates", "overlaps", "coordinates"))
```

## 3. Visualise receiver coverage

`plotDeployments()` is the visual companion to the audit — a timeline of when each receiver was
deployed and actively listening. It reads the same log, breaks a bar wherever coverage has a real
gap, and (invisibly) returns a per-receiver coverage table (active days, gaps, coverage fraction).

```{r}
#| eval: false
plotDeployments(deployments)
```

With a richer log — receivers moved between stations, servicing gaps, staggered deployments — group
the rows by a coarser unit and colour them with `group.by`, e.g. `plotDeployments(dep, row.by =
"station", group.by = "array")`.

## 4. Match detections to deployments

```{r}
#| eval: false
detections <- matchDeployments(rays_detections, deployments,
                               station.col = "station")
```

Then attach the tag metadata, exactly as in [module 01](01-import.html). This is what makes the next
step short: `assignAnimalIDs()` derives each animal's tagging date and transmitter nominal delay and
stores them on the object, so `filterDetections()` does not have to be told them.

```{r}
#| eval: false
tags <- importTags(rays_tags, source = "generic",
                   col.map = list(ID = "ID", tagging_date = "tagging_date"))
detections <- assignAnimalIDs(detections, tags)

mobyMeta(detections)$tagging.dates   # already attached
```

## 5. Filter spurious detections

`filterDetections()` applies a sequence of opt-in quality-control criteria: exact-duplicate removal,
detections before tagging or after a cut-off, a per-receiver short-interval **false-detection filter**
(`nominal.delay`, scaled to the transmitter delay — the standard code-collision filter of Pincock 2012 /
Simpfendorfer et al. 2015), an optional residency/isolation filter (`isolation.window`, off by default),
and an **unrealistic-speed** filter. Removed detections are returned in `data_discarded` with a `reason`;
nothing is deleted silently.

Note what is *not* in the call below: no column arguments and no tagging dates — they travel with the
object, so only the filter settings themselves need naming.

```{r}
#| eval: false
filtered <- filterDetections(
  detections,
  max.speed  = 2,            # m/s; great-circle distances (no land.shape needed)
  speed.unit = "m/s"
)
filtered                        # a printable <mobyFilter> summary
clean <- filtered$data          # the cleaned detections (with a qc_flag column)
```

::: {.callout-note}
The short-interval false-detection filter needs a transmitter nominal delay, which this example tag
table does not carry — so it stays off here. When your tag export has a delay (or a min/max delay
range), `importTags()` picks it up and `assignAnimalIDs()` attaches it per animal, and the filter then
runs with no extra arguments. You can also pass `nominal.delay` directly, or set a fixed per-receiver
window with `min.lag.threshold` when the delay is unknown. See `?filterDetections`, section
*Isolation-based filtering*.
:::

## 6. Bin detections into regular intervals

Most downstream analyses operate on regular **time bins** rather than raw detections.

```{r}
#| eval: false
clean$timebin <- getTimeBins(clean$datetime, interval = "1 hour")
```

::: {.callout-tip}
The result of this module is equivalent to the bundled `rays` checkpoint, which later modules load
directly so you can start anywhere.
:::

## Recap & what's next

You audited, matched, filtered and binned the data into an analysis-ready `mobyData`.

➡️ **Next:** [03 — Exploratory analysis](03-exploratory.html).
