---
title: "03 - Exploratory analysis"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 9, fig.height = 5.5, dpi = 96, fig.align = "center")
```

::: {.callout-note}
## In this module
Starting from a clean `mobyData` checkpoint, you will visualise detection patterns, classify diel
phases, quantify residency, and produce a publication-style summary table — **separately for each
species** using a single `id.groups` object.

**Prerequisites:** none. This module loads the bundled checkpoint `rays`, so you can run it without
having completed modules 01–02.
:::

## Learning objectives

By the end of this module you will be able to:

1. Load the cleaned `mobyData` checkpoint and inspect its metadata.
2. Use **`id.groups`** to generate every output per species automatically.
3. Draw abacus, station-summary and chronogram plots.
4. Classify detections by **diel phase** with `getDielPhase()`.
5. Reveal activity **rhythms** with `plotPeriodogram()` and `plotScalogram()`.
6. Compute **residency indices** with `calculateResidency()` and format them with `summaryTable()`.

## Setup

```{r}
#| message: false
library(moby)

# the cleaned, analysis-ready checkpoint (a mobyData object: see ?rays)
data(rays)

# the two-species grouping is stored in the object's metadata
id_groups <- mobyMeta(rays)$id.groups
names(id_groups)
```

::: {.callout-tip}
## The `id.groups` motif
`id_groups` is a **named list of animal IDs, one element per species**. Passing it to a moby
function tells that function to compute and lay out its results *independently for each group*. You
will pass the same `id_groups` object to almost every function from here on — that is all it takes
to get per-species outputs.
:::

```{r}
str(id_groups)
```

## 1. A first look: the abacus plot

An *abacus plot* shows, for every individual, when it was detected over the study period — the
quickest way to eyeshot data coverage, gaps and tag attrition.

```{r}
#| fig-height: 6
plotAbacus(rays, id.groups = id_groups)
```

## 2. Where were animals detected? Station summaries

Each statistic is drawn in its own panel with its own axis, so counts of detections and of
individuals are never conflated on a shared scale. Columns split the animals by species.

```{r}
#| fig-height: 6
plotStationStats(rays, type = c("detections", "individuals"), id.groups = id_groups)
```

## 3. Diel patterns

Many coastal species shift activity between day and night. First we classify every detection by
**diel phase** (here a simple day/night split based on the array's location), then visualise the
hour-by-date structure with a chronogram.

```{r}
# representative coordinates of the array (centre of the MPA)
mpa_coords <- matrix(c(mean(rays$lon), mean(rays$lat)), nrow = 1)

rays$diel <- getDielPhase(rays$datetime, coords = mpa_coords, phases = 2)
table(rays$diel)
```

```{r}
#| fig-height: 6
# chronograms operate on time-binned detections; bin to the hour first
rays$timebin <- getTimeBins(rays$datetime, interval = "60 mins")

# one panel per species (split.by), points coloured by receiver
plotChronogram(rays, coords = mpa_coords, split.by = "species",
               color.by = "station")
```

## 4. Activity rhythms

The chronogram *shows* diel structure; `plotPeriodogram()` and `plotScalogram()` *quantify* the
periodicities in a detection time series. Both work on a binned count series, so first collapse the
detections into hourly counts per individual (`rays$timebin` was created above).

```{r}
# hourly detection counts per individual: a long ID x time-bin x count table
binned <- aggregate(list(detections = rep(1L, nrow(rays))),
                    by = list(ID = rays$ID, timebin = rays$timebin), FUN = sum)
```

A **periodogram** gives one global spectrum per animal — how much power sits at each candidate
period. The tidal (~6–12 h) and diel (~24 h) bands are shaded and the dominant peak is marked.

```{r}
#| fig-height: 6.5
plotPeriodogram(binned, id.col = "ID", timebin.col = "timebin", id.groups = id_groups)
```

A **scalogram** (continuous wavelet transform) shows *when* each rhythm is present — a time (x) by
period (y) heat map. Shown here for the first species:

```{r}
#| fig-height: 6
plotScalogram(binned[binned$ID %in% id_groups[[1]], ],
              variable = "detections", id.col = "ID", timebin.col = "timebin", ncol = 2)
```

::: {.callout-note}
For these simulated rays the strongest periodicities are multi-day presence bouts, with a weaker
near-diel (~24 h) component. Real activity data with a clear circadian rhythm shows a sharp 24 h
peak and a persistent diel band running across the scalogram.
:::

## 5. Residency

`calculateResidency()` returns a tidy, numeric table of residency indices (IR1, IR2, IWR) and the
building blocks they derive from. We supply the date monitoring ended so the indices that need the
full study interval can be computed.

```{r}
monitoring_end <- as.POSIXct("2023-07-05", tz = "UTC")

residency <- calculateResidency(rays, last.monitoring.date = monitoring_end)
head(residency)
```

## 6. A publication-ready summary

`summaryTable()` formats the residency values (and detection/coverage metrics) into a
per-individual table, grouped by species when `id.groups` is supplied.

```{r}
summaryTable(rays, last.monitoring.date = monitoring_end, id.groups = id_groups)
```

## Recap & what's next

You loaded a checkpoint, drove every output per species with one `id.groups` object, and moved from
raw detections to residency indices and a formatted summary.

➡️ **Next:** [04 — Movement & space use](04-space-use.html), where we turn detections into positions,
distances and home ranges.
