---
title: "04 - Movement & space use"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 9, fig.height = 6, dpi = 96, fig.align = "center")
```

::: {.callout-note}
## In this module
Turn detections into positions, then quantify how much animals move and the areas they use: centres
of activity, step distances, rates of movement, and home ranges (kernel utilisation distributions),
all per species.

**Prerequisites:** none. Loads the `rays` checkpoint.
:::

## Learning objectives

1. Estimate positions (**centres of activity**) with `calculateCOAs()`.
2. Reconstruct movement paths and step distances with `calculateStepDistances()`.
3. Summarise movement with `calculateROM()` and `calculateLinearityIndex()`.
4. Estimate home ranges with `calculateUDs()` (autocorrelated KDE) and map them with `plotMaps()`.

## Setup

```{r}
#| message: false
library(moby)
data(rays)
id_groups <- mobyMeta(rays)$id.groups
```

## 1. Centres of activity (positions)

Acoustic detections are presence-at-a-receiver events; `calculateCOAs()` converts them into
short-interval position estimates suitable for movement and home-range analysis.

```{r}
#| eval: false
coas <- calculateCOAs(rays)   # aggregates the detections in each time bin (rays already carries 'timebin')
```

## 2. Step distances and movement tracks

```{r}
#| eval: false
# great-circle step distances (supply a land.shape for shortest in-water paths)
tracks <- calculateStepDistances(coas)
getTrajectories(tracks)   # the reconstructed path geometries (an attribute)
```

## 3. Rates of movement & directness

```{r}
#| eval: false
rom <- calculateROM(tracks)                 # total distance + mean/max rate of movement
li  <- calculateLinearityIndex(tracks)      # movement directness (0 = convoluted, 1 = straight)
```

## 4. Home ranges (UDs) and maps

`calculateUDs()` defaults to **autocorrelated kernel density estimation (AKDE)** via `ctmm`, which
accounts for the autocorrelation inherent in tracking data.

```{r}
#| eval: false
kud <- calculateUDs(coas, id.groups = id_groups)   # 'coas' carries the CRS forward from 'rays'

plotMaps(coas, uds = kud, animal.tracks = tracks,
         id.groups = id_groups)
```

## 5. A movement summary table

```{r}
#| eval: false
movementTable(tracks, ud.results = kud, id.groups = id_groups)
```

## Recap & what's next

You estimated positions, movement metrics and home ranges per species.

➡️ **Next:** [05 — Movement (spatial) networks](05-movement-networks.html).
