---
title: "05 - Movement (spatial) networks"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 9, fig.height = 6, dpi = 96, fig.align = "center")
```

::: {.callout-note}
## In this module
Build a **movement network** where nodes are locations (receivers/sites) and directed edges are
transitions between them. moby treats the two network types — movement and association — symmetrically,
so the workflow here is the same four-verb rhythm you will reuse in module 06:
**calculate → metrics → randomise → plot**.

**Prerequisites:** none. Loads the `rays` checkpoint.
:::

## Learning objectives

1. Build a movement network with `calculateTransitions()`.
2. Quantify it with `networkMetrics()`.
3. Test it against a null model with `randomizeTransitions()`.
4. Map it with `plotMovements()` and tabulate it with `transitionsTable()`.

## Setup

```{r}
#| message: false
library(moby)
data(rays)
id_groups <- mobyMeta(rays)$id.groups
```

## 1. Build the network

A transition is a movement between two *different* consecutive **residence events** (visits) in each
animal's time-ordered sequence. Visits are segmented by `max.gap` (default 48 h): consecutive
detections at the same station are one visit unless separated by a longer absence, in which case the
later return is treated as a new visit rather than one continuous stay. Supplying `id.groups` builds
one network per species.

```{r}
#| eval: false
mov_net <- calculateTransitions(rays, spatial.col = "station", id.groups = id_groups)
mov_net                       # a mobyNetwork object (edge list + node table)
```

The node table reports `n_residence` (the number of visits to each station) and `mean_residence_h`
(mean visit duration); edge `mean_duration_h` is the mean transit time over continuously-observed
moves only (transits that spanned a `> max.gap` absence are kept for connectivity but flagged
`crossed_gap` and excluded from that mean).

::: {.callout-tip}
If you want the discrete visits themselves — one row per arrival/departure — rather than the network
summary, call `calculateVisits(rays, spatial.col = "station")`. It shares the same `max.gap`
segmentation, so its counts and durations line up with the network's node table.
:::

## 2. Network metrics

```{r}
#| eval: false
networkMetrics(mov_net)       # node- and network-level graph metrics
```

## 3. Null model

```{r}
#| eval: false
mov_rand <- randomizeTransitions(mov_net, iterations = 1000)
```

## 4. Visualise & tabulate

```{r}
#| eval: false
plotMovements(mov_net, id.groups = id_groups)   # also: plot(mov_net)
transitionsTable(mov_net)
```

::: {.callout-tip}
`plot()` on any `mobyNetwork` dispatches to the right routine automatically — `plotMovements()` for
movement networks, `plotAssociations()` for association networks.
:::

## Recap & what's next

You built, measured, tested and mapped a movement network per species.

➡️ **Next:** [06 — Association (social) networks](06-association-networks.html) — the same four steps,
applied to co-occurrence between individuals.
