---
title: "06 - Association (social) networks"
---

```{r}
#| include: false
#| purl: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 9, fig.height = 6, dpi = 96, fig.align = "center")
```

::: {.callout-note}
## In this module
Build an **association network** where nodes are individuals and edges are co-occurrence-based
association indices (how often two animals are detected together in space and time). The workflow is
deliberately identical to module 05 — **calculate → metrics → randomise → plot** — over the *same*
`mobyNetwork` object and `networkMetrics()` function.

**Prerequisites:** none. Loads the `rays` checkpoint.
:::

::: {.callout-warning}
## Interpretation
Co-occurrence indices (SRI/HWI) measure *shared space-and-time use*, **not** confirmed social
interaction. Always pair them with a randomisation test, and report associations with
multiple-comparison control.
:::

## Learning objectives

1. Build an association network with `calculateAssociations()`.
2. Quantify it with `networkMetrics()` (same function as module 05).
3. Test it against a datastream-permutation null with `randomizeAssociations()`.
4. Visualise it with `plotAssociations()`, `plotAssociationMatrix()` and `plotGroupSizeDistribution()`.

## Setup

```{r}
#| message: false
library(moby)
data(rays)
id_groups <- mobyMeta(rays)$id.groups
```

## 1. Build the network

Associations are usually computed *within* species; passing `id.groups` builds one network per group.

```{r}
#| eval: false
rays_wide <- createWideTable(rays, value.col = "station")   # time-bin x individual co-occurrence table
assoc_net <- calculateAssociations(rays_wide, id.groups = id_groups)
assoc_net                       # a mobyNetwork object
```

## 2. Network metrics

```{r}
#| eval: false
networkMetrics(assoc_net)
```

## 3. Null model (with multiple-comparison control)

```{r}
#| eval: false
assoc_rand <- randomizeAssociations(rays_wide, assoc_net, iterations = 1000,
                                    p.adjust.method = "fdr")
```

## 4. Visualise

```{r}
#| eval: false
plotAssociations(assoc_net)        # also: plot(assoc_net)
plotAssociationMatrix(assoc_rand)  # the matrix needs the null-model (randomised) result
plotGroupSizeDistribution(rays, id.groups = id_groups)
```

## Recap

You have now run both moby network pipelines and seen how the shared `mobyNetwork` object and
`networkMetrics()` make movement and association analyses mirror each other.

🎉 That completes the core tutorial series. For pulling your own data from a live database, see the
optional [ETN appendix](00-etn-import.html).
